Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2016-09-01 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-perf: tidy-up discussion of GPU results.

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to a shallow-water model. We found that although
the re-structuring did initially incur a sizeable performance penalty,
it was possible to transform the resulting code to recover performance
while obeying the \psykal\ separation of concerns. In this work we
move to looking at parallel performance within the \psykal\ approach.

@@

In common with OP2, the \psykal\ approach separates out the science
code and the performance-related code into distinct layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics. However, the
\psykal\ approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. Another difference is that the description
metadata is provided with a kernel in the \psykal\ approach whereas it
is provided as part of the kernel call in OP2.

@@

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on fields
that are local to the process doing the computation. (Depending on the
type of kernel, these may be a set of elements, a single column of
elements, or a set of columns.). Again the scientist is responsible
for writing this layer and there is no parallelism specified here,
but, depending on the complexity of the Kernels, there may be input
from an HPC expert and/or some coding rules to help ensure the kernels
compile into efficient code.  In an alternative approach, kernels are
generated from a high-level specification~\cite{firedrake,fenics},
potentially allowing them to be optimised automatically~\cite{coffee}.

@@

As the PSy layer iterates over columns, the single core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPGPUs, or some combination thereof) and software (such as
compiler, operating system, MPI implementation) configuration with no
change to the Algorithm or Kernel layer code. This approach therefore
offers the potential for portable performance. In this work we apply
optimisations to the PSy layer manually. The development of a tool to
automate this process will be the subject of a future paper.

Clearly the separation of code into distinct layers may have an effect
on performance. This overhead, how to get back to the performance
of a hand-optimised code, and potentially improve on it, will be
discussed in the remainder of this paper.

@@

The order of the model computation is as follows:
\begin{enumerate}
\item Set the initial conditions (water depth, sea surface height, velocity)
\item integrate the continuity equation for the new sea surface height
\item update the different terms in the right hand side of the momentum equations; advection, Coriolis forcing (if set in explicit form), pressure gradient, and horizontal viscosity
\item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
\item apply the boundary conditions on the open-  and solid-boundary cells.
\end{enumerate}

@@

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop around each kernel call.

@@

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
%3.7~GHz and 10~MB of last-level cache. 
The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB.

@@

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity has a loop-carried dependency
in $j$ which appears to prevent its being executed in
parallel.\footnote{Only once this work was complete did we establish
  that boundary conditions are enforced such that it can be executed
  in parallel.}  This was therefore left unchanged and executed on
thread 0 only.

@@

Having show that we can recover, and often improve upon, the
performance of the original version of NEMOLite2D, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation has been
applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256} where the performance of the \psykal\
version for a given compiler/CPU is given relative to the performance
of the original with the same compiler/CPU.

@@

Since the space to explore consists of three different compilers, six
different domain sizes, five stages of optimisation and six different
thread counts we can only consider a slice through it in what follows.
In order to inform our choice of domain size,
Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
most performant, Intel-compiled version of NEMOLite2D. Although it is
common for production runs of NEMO to use MPI sub-domains of $20
\times 20$ ({\bf do we have/need a reference for that?}), it is clear
from Figure~\ref{FIG_omp_domain_size} that the limited quantity of
parallelism in the $32 \times 32$ domain inhibits scaling. Therefore,
to fully test the performance of our OpenMP implementations we have
chosen to examine results for the $256 \times 256$ domain and these
are shown in Figures~\ref{FIG_gnu_omp_256}, \ref{FIG_intel_omp_256}
and \ref{FIG_cray_omp_256}.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal\
  NEMOLite2D for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

The simplest OpenMP implementation (black lines, circle symbols) fails
to scale well for any of the compilers. For the Intel and Gnu
versions, parallel efficiency is already less than 50\% on just four
threads. The Cray version however does better and is about 45\%
efficient on eight threads.

@@

We have used different scales for the $y$-axes in each of the plots in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} in order to highlight the performance
differences between the code versions with a given compiler. However,
the best performance obtained on twelve threads ({\it i.e.}  without
hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
Intel and Cray compilers, respectively. This serves to emphasise the
democratisation that the introduction of OpenMP has had; for the
serial case the Cray- and Intel-compiled exectables were a factor of
two faster than the Gnu-compiled binary
(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
Gnu-compiled binary is only 6\% slower than that produced by the Intel
compiler and is 13\% {\em faster} that that of the Cray compiler. In
part this situation comes about because of the effect that adding the
OpenMP compiler flag has on the optimisations performed by the
compiler.  In particular, the Cray compiler no-longer vectorises the
key Momentum section, despite the directive added during the serial
optimisation work.

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
configuration we ensured that threads were evenly shared over the two
sockets. The performance obtained for the $256^2$ case with the `early
SINGLE' version of the code is compared with the single-socket
performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
available memory bandwidth in this way has little effect on
performance - the two-socket performance figures track those from a
single socket very closely. The only significant difference in
performance is at 24 threads where, in addition to the difference in
available memory bandwidth, the single-socket configuration is using
hyperthreading while the two-socket case is not.  The discrepancy in
the performance of the Cray- and Intel-compiled binaries at this
thread count is under investigation by Cray.

@@

The introduction of OpenACC directives proved to be simpler than
producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
  whether this is true}.  Consequently we only show the performance of
{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the various GPU implementations of NEMOLite2D.}
\label{FIG_gpu_opt_stages}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal\
  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
\label{FIG_cpu_cf_gpu}
\end{figure}

Once the problem size is increased to $128^2$, a single time-step
takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
comparable in performance to the OpenACC version. For all of the
larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
version is considerably faster than the CPU.  For problem sizes of
$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
exhausted and performance becomes limited by the bandwidth to main
memory. At this stage the OpenACC version of the code on the GPU is
some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
than the OpenMP version on the CPU. {\bf Can we relate this to the
  difference in peak memory b/w between the GPU and CPU?}

@@

Although we have investigated how the performance of the \psykal
version of NEMOLite2D compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{roofline}
\caption{Comparison of the performance achieved by NEMOLite2D and
  Shallow on the E5-1620 CPU. Results for the former are for the
  $256^2$ domain since that gave the best performance.}
\label{FIG_roofline}
\end{figure}

In Figure~\ref{FIG_roofline} we show the performance of kernels from
both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
This demonstrates that the Shallow kernel is achieving a performance
roughly consistent with saturating the available memory bandwidth. In
contrast, the kernel taken from NEMOLite2D is struggling to reach a
performance consistent with saturating the bandwidth to main
memory. We experimented with reducing the problem size (so as to
ensure it fitted within cache) but that did not significantly improve
the performance of the kernel. This then points to more fundamental issues
with the way that the kernel is implemented which are not captured in the
simple roofline model.

In order to aid our understanding of kernel performance we have
developed a tool capable of parsing Fortran code and generating a
Directed Acyclic Graph (DAG) of the data flow. This tool eases the
laborious and error-prone process of counting memory accesses and
FLOPs as well as providing information on those operations that are
rate-limiting or on the critical path. Using the details of the Intel
Ivy Bridge microarchitecture published by Fog~\cite{fog_microarch,
  fog_instructions} we have constructed performance estimates of the
NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
(i.e. assuming that all nodes in the DAG are executed in serial) we
get a lower-bound performance estimate of $0.6391 \times CLOCK\_SPEED$
FLOPS which gives 2.46~GFLOPS at a clock speed of
3.85~GHz.

Alternatively, we may construct an upper bound by assuming the
out-of-order execution engine of the Ivy Bridge core is able to
perfectly schedule and pipeline all operations such that those that go
to different execution ports are always run in parallel. In the Ivy
Bridge core floating-point multiplication and division operations go
to port 0 while addition and subtraction go to port
1~\cite{fog_instructions}. Therefore we sum the cost of all
multiplication and division operations in the DAG and compare that
with the sum of all addition and subtraction operations. The greater
of these two quantities is then taken to be the cost of executing the
kernel; all of the operations on the other port are assumed to be done
in parallel. This gives a performance estimate of $1.029 \times
CLOCK\_SPEED$ FLOPS or 3.96 GFLOPS at 3.85~GHz.

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is clearly sub-optimal, this is due to the
balance of floating-point operations that it contains. Improving this
is outside the scope of the current work since here we are focused on
the \psykal separation of concerns and the introduction of parallelism
in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE by assuming perfect vectorisation which gives a
performance increase of a factor of two.

@@

We have investigated the application of the \psykal\ separation of
concerns approach to the domain of shared-memory parallel,
finite-difference ocean models. This approach enables the
computational science (performance) related aspects of a computer
model to be kept separate from the natural (oceanographic) science
aspects.

We have used a new and un-optimised, two-dimension model extracted
from the NEMO ocean model for this work. As a consequence of this, the
introduction of the \psykal\ separation of concerns followed by
suitable transformations of the PSy Layer is found to improve
performance. This is in contrast to our previous
experience~\cite{shallow_psykal} with tackling the Shallow code which
has had been optimised over the course of several years. In that case
we were able to recover\ldots

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal\ version of a code.

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}) and some {\bf NVIDIA system - or is it
  Emerald?}.

>>>>

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model. We found
that although the re-structuring did initially incur a sizeable
performance penalty, it was possible to transform the resulting code
to recover performance (for a variety of CPU/compiler combinations)
while obeying the \psykal\ separation of concerns. In this work we
move to looking at portable parallel performance within the
\psykal\ approach.

@@

In common with OP2, the \psykal\ approach separates out the science
code and the performance-related code into distinct layers. The calls
that specify parallelism in both approaches are similar in terms of
where they are placed in the code and in their semantics. However, the
\psykal\ approach supports the specification of more than one kernel
in a parallel region of code, compared with one for OP2, giving more
scope for optimisation. In addition, the metadata describing a kernel
is included with the kernel code in the \psykal\ approach whereas it
is provided as a part of the kernel call in OP2.

@@

The bottom layer, in terms of calling hierarchy, is the Kernel
layer. The Kernel layer implements the science that the Algorithm
layer calls, as a set of subroutines. These kernels operate on fields
that are local to the process doing the computation. (Depending on the
type of kernel, these may be a set of elements, a single column of
elements, or a set of columns.) Again the scientist is responsible for
writing this layer and there is no parallelism specified here, but,
depending on the complexity of the Kernels, there may be input from an
HPC expert and/or some coding rules to help ensure that the kernels
compile into efficient code.  In an alternative
approach~\cite{firedrake,fenics}, kernels are generated from a
high-level specification, potentially allowing them to be optimised
automatically~\cite{coffee}.

@@

As the PSy layer iterates over columns, the single-core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPGPUs, or some combination thereof) and software (such as
compiler, operating system, MPI implementation) configuration with no
change to the Algorithm or Kernel layer code. This approach therefore
offers the potential for portable performance. In this work we apply
optimisations to the PSy layer manually. The development of a tool to
automate this process will be the subject of a future paper.

Clearly the separation of code into distinct layers may have an effect
on performance. This overhead, how to get back to the performance of a
parallel, hand-optimised code, and potentially improve on it, will be
discussed in the remainder of this paper.

@@

The sequence of the model computation is as follows:
\begin{enumerate}
\item Set the initial conditions (water depth, sea surface height, velocity)
\item integrate the continuity equation for the new sea surface height
\item update the different terms in the right hand side of the momentum equations; advection, Coriolis forcing (if set in explicit form), pressure gradient, and horizontal viscosity
\item update the velocity vectors by summing up the values in 3), and implicitly with the bottom friction and Coriolis forcing (if set in implicit form)
\item apply the boundary conditions on the open- and solid-boundary cells.
\end{enumerate}

@@

In order to maximise the flexibility (and thus potential for
optimisation) of the \psykal\ version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop (over the two dimensions of
the model grid) around each kernel call.

@@

Our aim in this work is to achieve portable performance, especially
between multi-core CPU and many-core GPU systems. Consequently, we
have performed tests on both an Intel Ivy Bridge CPU (E5-2697 at
2.7~GHz) and on an NVIDIA Tesla K40 GPU.  On the Intel-based system we
have used the Gnu, Intel and Cray Fortran compilers (versions 4.9.1,
15.0.0.090 and 8.3.3, respectively). The code that made use of the GPU
was compiled using version 15.5 of the PGI compiler.

@@

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity ($v$) has a loop-carried
dependency in $j$ which appears to prevent its being executed in
parallel.\footnote{Only once this work was complete did we establish
  that boundary conditions are enforced such that it can safely be
  executed in parallel.}  This was therefore left unchanged and
executed on thread 0 only.

@@

Having show that we can recover, and often improve upon, the
performance of the original version of NEMOLite2D, the next logical
step is to examine the necessary code transformations in detail.  We
do this for the $256^{2}$ case since this fits within cache on the Ivy
Bridge CPUs we are using here.  Table~\ref{TABLE_opt_breakdown} shows
detailed performance figures for this case after each transformation
has been applied to the code. The same data is visualised in
Figure~\ref{FIG_opt_stages_256}.

@@

Since the space to explore consists of three different compilers, six
different domain sizes, five stages of optimisation and six different
thread counts we can only consider a slice through it in what follows.
In order to inform our choice of domain size,
Figure~\ref{FIG_omp_domain_size} shows the scaling behaviour of the
most performant, Intel-compiled version of NEMOLite2D. Although it is
common for production runs of NEMO to use MPI sub-domains of $20
\times 20$, it is clear from Figure~\ref{FIG_omp_domain_size} that the
limited quantity of parallelism in the $32 \times 32$ domain inhibits
scaling. Therefore, to fully test the performance of our OpenMP
implementations we have chosen to examine results for the $256 \times
256$ domain and these are shown in Figures~\ref{FIG_gnu_omp_256},
\ref{FIG_intel_omp_256} and \ref{FIG_cray_omp_256}.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of
  \psykal\ NEMOLite2D for the $256^2$ domain with the Gnu compiler on
  a single Intel Ivy Bridge socket. The corresponding parallel
  efficiencies are shown using open symbols and dashed lines. The
  24-thread runs employed hyperthreading and the optimal OpenMP
  schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

The simplest OpenMP implementation (black lines, circle symbols) fails
to scale well for any of the compilers. For the Intel and Gnu
versions, parallel efficiency is already less than 50\% on just four
threads. The Cray version however does better and is about 45\%
efficient on eight threads (Figure~\ref{FIG_cray_omp_256}).

@@

We have used different scales for the $y$-axes in each of the plots in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} in order to highlight the performance
differences between the code versions with a given compiler. However,
the best performance obtained on twelve threads ({\it i.e.}  without
hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
Intel and Cray compilers, respectively. This serves to emphasise the
democratisation that the introduction of OpenMP has had; for the
serial case the Cray- and Intel-compiled exectables were a factor of
two faster than the Gnu-compiled binary
(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
Gnu-compiled binary is only 6\% slower than that produced by the Intel
compiler and is 13\% {\em faster} that that of the Cray compiler. In
part this situation comes about because of the effect that adding the
OpenMP compiler flag has on the optimisations performed by the
compiler.  In particular, the Cray compiler no longer vectorises the
key Momentum section, despite the directive added during the serial
optimisation work. This defficiency has been reported to Cray.

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of ARCHER, a Cray
XC30). For this configuration we ensured that threads were evenly
shared over the two sockets. The performance obtained for the $256^2$
case with the `early SINGLE' version of the code is compared with the
single-socket performance in
Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the available
memory bandwidth in this way has little effect on performance - the
two-socket performance figures track those from a single socket very
closely. The only significant difference in performance is at 24
threads where, in addition to the difference in available memory
bandwidth, the single-socket configuration is using hyperthreading
while the two-socket case is not.  The discrepancy in the performance
of the Cray- and Intel-compiled binaries at this thread count is under
investigation by Cray.

@@

The advantage of the \psykal\ re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the \psykal\ re-structuring, we
had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the \psykal
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the \psykal/OpenACC approach.

The performance of these two versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  Since the \psykal restructuring of
NEMOLite2D did not prove to be a barrier to achieving a performant
code when using OpenACC directives, we only report the performance of
the \psykal\ OpenACC version. For problem sizes up to $256^2$ the
difference in the performance of the two versions is less than 5\% and
for the larger domains it is at most 11\%. This demonstrates that the
performance cost of the \psykal\ approach in utilising a GPU for
NEMOLite2D is minimal.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the \psykal\ (OpenACC) and CUDA GPU
  implementations of NEMOLite2D. The performance for the OpenACC
  version of the original code is not shown since it is identical to
  that of the \psykal\ version.}
\label{FIG_gpu_opt_stages}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal\
  NEMOLite2D (on a single Intel Ivy Bridge socket) compared with the \psykal\ GPU implementation (using OpenACC).}
\label{FIG_cpu_cf_gpu}
\end{figure}

Once the problem size is increased to $128^2$, a single time-step
takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
comparable in performance to the OpenACC version. For all of the
larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
version is considerably faster than the CPU.  For problem sizes of
$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
exhausted and performance becomes limited by the bandwidth to main
memory. At this stage the OpenACC version of the code on the GPU is
some 3.4 times faster than the best OpenMP version on the CPU.

%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
%memory, respectively on the CPU.

@@

Although we have investigated how the performance of the \psykal\
version of NEMOLite2D compares with that or the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\cite{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\cite{stream} and LINPACK~\cite{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{roofline}
\caption{Comparison of the performance achieved by kernels from
  NEMOLite2D and Shallow on a roof-line plot for the E5-1620
  CPU. Results for the former are for the $256^2$ domain since that
  gave the best performance. See the text for a discussion of the
  different CPU ceilings (dashed lines).}
\label{FIG_roofline}
\end{figure}

In Figure~\ref{FIG_roofline} we show the performance of kernels from
both Shallow and NEMOLite2D on the roofline model for an Intel E5-1620 CPU.
This demonstrates that the Shallow kernel is achieving a performance
roughly consistent with saturating the available memory bandwidth. In
contrast, the kernel taken from NEMOLite2D is struggling to reach a
performance consistent with saturating the bandwidth to main
memory. We experimented with reducing the problem size (so as to
ensure it fitted within cache) but that did not significantly improve
the performance of the kernel. This then points to more fundamental issues
with the way that the kernel is implemented which are not captured in the
simple roofline model.

In order to aid our understanding of kernel performance we have
developed a tool capable of parsing Fortran code and generating a
Directed Acyclic Graph (DAG) of the data flow. This tool eases the
laborious and error-prone process of counting memory accesses and
FLOPs as well as providing information on those operations that are
rate-limiting or on the critical path. Using the details of the Intel
Ivy Bridge microarchitecture published by Fog~\cite{fog_microarch,
  fog_instructions} we have constructed performance estimates of the
NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
(ILP), \textit{i.e.} assuming that all nodes in the DAG are executed
in serial, we get a lower-bound performance estimate of $0.6391 \times
CLOCK\_SPEED$ FLOPS which gives 2.46~GFLOPS at a clock speed of
3.85~GHz.

Alternatively, we may construct an upper bound by assuming the
out-of-order execution engine of the Ivy Bridge core is able to
perfectly schedule and pipeline all operations such that those that go
to different execution ports are always run in parallel. In the Ivy
Bridge core, floating-point multiplication and division operations go
to port 0 while addition and subtraction go to port
1~\cite{fog_instructions}. Therefore we sum the cost of all
multiplication and division operations in the DAG and compare that
with the sum of all addition and subtraction operations. The greater
of these two quantities is then taken to be the cost of executing the
kernel; all of the operations on the other port are assumed to be done
in parallel. This gives a performance estimate of $1.029 \times
CLOCK\_SPEED$ FLOPS or 3.96 GFLOPS at 3.85~GHz.

These performance estimates are plotted as CPU ceilings (horizontal
dashed lines) in Figure~\ref{FIG_roofline}. The performance of the
Momentum kernel is seen to fall between these two bounds which
demonstrates that its performance is not memory-bandwidth limited, as
might have been assumed by its low AI. Therefore, although the
performance of this kernel is clearly sub-optimal, this is due to the
balance of floating-point operations that it contains. Improving this
is outside the scope of the current work since here we are focused on
the \psykal\ separation of concerns and the introduction of parallelism
in the PSy layer.

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE into that estimate by simply assuming perfect
vectorisation which gives a performance increase of a factor of two.

@@

We have investigated the application of the \psykal\ separation of
concerns approach to the domain of shared-memory parallel,
finite-difference shallow-water models. This approach enables the
computational science (performance) related aspects of a computer
model to be kept separate from the natural (oceanographic) science
aspects.

We have used a new and un-optimised, two-dimension model extracted
from the NEMO ocean model for this work. As a consequence of this, the
introduction of the \psykal\ separation of concerns followed by
suitable transformations of the PSy Layer is actually found to improve
performance. This is in contrast to our previous
experience~\cite{shallow_psykal} with tackling the Shallow code which
has had been optimised over the course of several years. In that case
we were able to recover more than 90\% of the performance of the
original for the Gnu, Intel, Cray and IBM compilers on both Intel and
IBM Power CPUs.

Investigation of the absolute serial performance of the NEMLite2D code
using the roofline model revealed that it was still significantly
below any of the traditional roofline ceilings. We have developed a
code-analysis tool~\cite{habakkuk} that is capable of providing more
accurate ceilings by analysing the nature of the floating point
computations performed by a kernel. In future work we aim to extend
this tool such that it can be applied to code parallelised using
OpenMP.

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal\ version of a code. For both
NEMLite2D and Shallow we have found that for serial performance, the
most important transformation is that of in-lining the kernel source
at the call site. However, for the more complex NEMOLite2D code the
Cray compiler also had to be coerced into performing SIMD
vectorisation through the use of souce-code directives.

In this work we have also demonstrated the introduction of parallelism
into the PSy layer with both OpenMP and OpenACC directives. In both
cases we were able to leave the natural-science parts of the code
unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
on 12 threads by enclosing the body of the (single) PSy layer routine
within a single PARALLEL region. Removal of unecessary synchronisation
points through use of the NOWAIT clause boosted 12-thread performance
by approximately 10\% with the Gnu and Cray compilers. The
\psykal\ re-structuring of this (admittedly small) code was not found
to pose any problems for the introduction of performant OpenMP.
Similarly, we have also demonstrated good GPU performance using OpenACC
in a \psykal\ version of the code.

Having demonstrated that the \psykal\ separation of concerns may be
applied to 2D finite-difference codes with minimal performance loss,
our next steps will be; 1) to consider the automatic generation of the
PSy layer and 2) to look at extending the approach to the full NEMO
model (i.e. three dimensions). In future work we will analyse the
performance of a domain-specific compiler that performs the automatic
generation of the PSy layer. This compiler (which we have named
`PSyclone') is currently under development.

@@

This work made use of the ARCHER UK National Supercomputing Service
(\url{http://www.archer.ac.uk}) and Emerald, a GPU-accelerated High
Performance Computer, made available by the Science \& Engineering
South Consortium operated in partnership with the STFC
Rutherford-Appleton Laboratory
(\url{http://www.ses.ac.uk/high-performance-computing/emerald/}).
